---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading the data
Loading data with standard function *read.csv*.
Loading library *Tidyverse* for operations with data frames and visualisations.

```{r, echo=TRUE,message= FALSE, results=FALSE}
library (tidyverse)
if (!exists("activity.csv")) unzip ("activity.zip")
data<-read.csv ("activity.csv")

```

## What is mean and median of total number of steps taken per day?
Proceeding data with function *summarise* and making histogram
```{r, echo=TRUE}

per_day<-summarise (group_by(data,date), day_steps=sum(steps,na.rm = TRUE))
print (paste("The mean is",mean(per_day$day_steps,na.rm = TRUE)))
print (paste("The median is",median(per_day$day_steps,na.rm=TRUE)))
```

# Histogram of the total number of steps taken each day
```{r}

pl1<-ggplot(per_day,aes(x=day_steps))
pl1+geom_histogram(breaks = seq(0, 25000, by=2500),col="red",fill="blue")+
  labs (title="Total number of steps taken each day", y="Frequency",
        x="Day steps")

```


## What is the average daily activity pattern?

# Plot of daily activity pattern
```{r}
per_interval<-summarise (group_by(data,interval), i_steps=mean(steps,na.rm = TRUE))
pl2<-ggplot(per_interval,aes(x=interval,y=i_steps))
pl2+geom_line()+labs (title="The average daily activity pattern", x= "Interval",
                      y="Average number of steps")
```

# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
print (paste ("Maximum activity interval - ", 
              per_interval[which.max(per_interval$i_steps), ]$interval))

```

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with \color{red}{\verb|NA|}NAs)
```{r}
print (sum(is.na(data$steps)))
```

Let`s use mean of 5-minute intervals for filling the missing values.  
Creating a new dataset **data2** with the missing data filled in.
```{r}
data2<-data
for (i in 1:nrow(data2))
{
  if (is.na(data2$steps[i])) 
    data2$steps[i]<-
      per_interval$i_steps[per_interval$interval==data2$interval[i]]
}
```
Comparing datasets with and without NAs

```{r}
per_day$type<-"With NAs"
per_day2<-summarise (group_by(data2,date), day_steps=sum(steps,na.rm = TRUE))
per_day2$type<-"Without NAs"
per_day3<-rbind (per_day,per_day2)

print (paste("The mean with NAs is",mean(per_day$day_steps,na.rm = TRUE)))
print (paste("The mean without NAs is",mean(per_day2$day_steps,na.rm = TRUE)))

print (paste("The median with NAs is",median(per_day$day_steps,na.rm=TRUE)))
print (paste("The median without NAs is",median(per_day2$day_steps,na.rm=TRUE)))
```

Let`s visualize the differences.
```{r}
pl3<-ggplot(per_day3,aes(x=day_steps))
pl3+geom_histogram(breaks = seq(0, 25000, by=2500),col="red",fill="blue")+
  facet_grid(.~type)+labs(title = "Total number of steps per day has some changes")
```


Because of chosen strategy of filling missing values daily activity pattern is not changed
```{r}
per_interval$type<-"With NAs"
per_interval2<-summarise (group_by(data2,interval), 
                          i_steps=mean(steps,na.rm = TRUE))
per_interval2$type<-"Without NAs"
per_interval3<-rbind(per_interval,per_interval2)

pl4<-ggplot(per_interval3,aes(x=interval,y=i_steps))
pl4+geom_line()+labs (title="The average daily activity pattern is not changes significantly", x= "Interval",
                      y="Average number of steps")+facet_grid(.~type)
```

## Are there differences in activity patterns between weekdays and weekends?

At first let`s add the new variable *weekday* to data frame *data2*.
By using function *weekdays* we define weekdays and weekends.
*Sys.setlocale* makes our script independent from local settings on current computer.
```{r, warning=FALSE,message=FALSE}
Sys.setlocale(category = "LC_ALL", locale = "English_United States.1252")
data2$date<-as.POSIXct(data2$date, "%Y%m%d")
data2$weekday<-"weekday"
data2$weekday[weekdays(data2$date)=="Saturday"|weekdays(data2$date)=="Sunday"]<-"weekend"
print (paste ("Mean of steps on weekdays is ", mean(data2$steps[data2$weekday=="weekday"])))

print (paste ("Mean of steps on weekends is ", mean(data2$steps[data2$weekday=="weekend"])))

```
Last step is making plot

```{r,warning=FALSE,message=FALSE}
per_interval_w<-summarise (group_by(data2,interval,weekday), 
                          i_steps=mean(steps,na.rm = TRUE))
pl5<-ggplot(per_interval_w,aes(x=interval,y=i_steps))
pl5+geom_line()+labs (title="The average daily activity pattern", x= "Interval",
                      y="Average number of steps")+facet_grid(weekday~.)
```


As we can see on weekends mean of activity is higher, but has more plain distribution

